Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
4.3

Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. Recall that when a
function is called in R, parameters are bound to unevaluated code thunks
called promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value
for type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \contractr modifies the promise
by wrapping the unevaluated promise expression in a call to its type-checker
along with the context necessary to associate the promise to the function
and its parameter (for failure messages). The contract checking happens when
the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \contractr receives values
that can be immediately type-checked. Non-strict semantics dictate that any
computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would
be incorrect to type-check it at the beginning of call. To make this work,
\contractr mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.

Secondly, we faced the problem of type-checking function return value. In R,
the result of a function call is the result of evaluating the last
expression in the function body; so an explicit return call is absent from
most function definitions. Thus, there could be many potential
sub-expressions in the function body which need to be type-checked against
the return type. To address this, \contractr registers the type-checker to
be called on the return value through a function exit hook. This hook is
executed in the function call's environment after it has executed. While
this works, we have to contend with yet another wrinkle, R interpreter can
perform longjumps in its C implementation, which causes active R function
calls on the stack to be discarded. When they are discarded, their exit
hooks are called, which in this case calls the registered type-checker. But,
these functions are in the middle of an active computation, so they don't
have a return value to type-check. \contractr deals with this problem by
allocating a unique sentinel object which serves as the return value for
calls that are discarded. The function call exit hook does not call the
type-checker if this unique value happens to be the return value of the
call.

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\paragraph{Checking the values} A number of properties can be checked
by a simple tag check, namely wether a value is a: null, environment, vector
or scalar. Other properties require an inspection of the contents of a
value: the absence of \NA or the type of the elements of list. Union types
require checking for each of the alternatives. We do not check higher-order
functions in this version of the tool.

When argument values are wrapped in promises (this is not always the case
due to compiler optimizations), in order to retain the non-strict semantics
of R, the expression held in the promise is wrapped in a call to the type
checker, and type checking is delayed until the promise is forced. This
leads to corner cases such that the type checking of a function may happen
after that function has returned.

Return values require care as well.  Functions return the last expression
they evaluate, thus to avoid having to analyze the code of the called
function the checker will register a callback on the exit hook. The hook is
executed in the function call's environment. Another wrinkle is due to
longjumps which causes active function calls on the stack to be
discarded. When they are discarded, their exit hooks are called.  But, these
function do not have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the
return value for calls that are discarded. The exit hook does not call the
type-checker if it see the sentinel.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
